---
title: "Introduction to R"
output:
  html_document:
    theme: "flatly"
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    # keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The goal for this handout is to familiarize ourselves with R, RStudio, and RMarkdown and to gain intuition about how coding can be used in the field of population ecology. We will work through this markdown file, running lines of code and writing our own code to practice basic skills in R. 

Here is a list of learning objectives that we will cover in this worksheet. We will use these skills throughout the course, so it will make your life easier if you get these down early. 

* Basics of troubleshooting (AKA how to use the internet)
* R Projects
* Performing basic mathematical operations
* Using functions
* Installing packages
* Generating data
* Reading in data
* Basics of data wrangling
* Using RMarkdown

## Troubleshooting

Likely the most important skill to have under your belt for coding in R is how to troubleshoot. Regardless of your experience level with R, you are bound to run into issues in the form of error messages or undesirable outputs. Luckily, the internet exists. Not only does the internet exist, but the internet contains, within it, practically endless resources regarding R. All one has to do is copy and paste an error message into a search engine or Google "[x topic] in r" and they're bound to find forums and blog posts dedicated to their issue. Further, within recent years, the development of large language models (LLMs) such as ChatGPT have dramatically changed how many people troubleshoot coding issues (see [this blog post](https://www.zdnet.com/article/how-to-use-chatgpt-to-write-code-what-it-does-well-and-what-it-doesnt/) for a description of how you might use ChatGPT to help you code). We heavily encourage the use of the internet, including LLMs, to aid in your coding for this class.

Additionally, we heavily encourage you to rely on your peers in this class, as well as us (Jeremy & Jeff). Sometimes, you may face specific challenges regarding assignments that are difficult to answer with the help of the internet. _Do not waste hours of time searching the internet for answers._ In these cases, asking peers in lab or on Canvas may be a great option. Alternatively, Jeremy & Jeff are available in or after lecture/lab times as well as during our office hours which are posted on Canvas. However, there is a respectful way to ask people for help with troubleshooting code and we ask that you try to adopt this approach when asking for help. Modified from [this guide on formatting questions on a popular coding forum](https://stackoverflow.com/help/how-to-ask), we request that you: 

- Clearly articulate the goal of your code, including the larger goal and any steps in the process.
- Describe what you have tried and how the output diverges from your desired or expected outcome. 
- Present the code that you have tried. 

If you are requesting help via e-mail, please make sure to follow these steps and supply your code in the form of a R Script or RMarkdown file. 

__A final, important note on code troubleshooting:__ To recieve help with code, you MUST know what you are trying to do. In this class, I will ask you to use code to construct and interrogate ecological models. To do this, you will need to understand the model and how you are trying to interrogate it. Many coding problems in this class can be addressed by first by asking yourself "Do I understand what I am trying to code?". If the answer is "no", you must resolve this before asking for help troubleshooting the code itself. Utilize the textbooks, the internet, peers, Jeremy or Jeff for understanding the conceptual components of assignments. 

## RStudio Projects

RStudio Projects are a way to manage and organize coding projects. If you work on multiple coding projects using RStudio, it is helpful to keep your workspaces and histories separate. Further, and perhaps most useful for you, is that working in an RStudio Project keeps consistent your working directory. We highly recommend creating an R Project for this class within some folder on you desktop. Use [this guide](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) to create an RStudio project in either a new directory on your computer or in an existing directory if you've already saved this markdown file to a particular directory. A third option exists for creating a directory that is linked to a repository on GitHub, which may or may not be of interest to you. 

## R Markdown

Markdown files are a compilation of text and lines of code used for 'literate programming'. The goal of markdown files aren't just to code up some process, but also to communicate the process and output of the code. You can imagine that reading a couple hundrer lines of R code might not be nearly as helpful as reading explanations of the problem, explanations of the approach, the code, figures produced by the code, and interpretation of the figures. 

There are lots of ways to customize R Markdown files, and there is always more to learn about them, but for now, we'll stick to the basics. What you're reading right now is written in markdown syntax, which is mostly just basic text with occasional fancy operators for formatting. We can put in R code by creating a "chunk". A chunk is initiated with three backtick symbols, followed by {r chunk_title}. A chunk is closed with another three backtick symbols. Check the .Rmd file to find examples of chunks. You can also click the "Insert a new code chunk" button while working in a .Rmd file, which should be the green box with a 'c' and a plus sign. 

While completing this worksheet, you will need to type into the markdown file itself ("Week One.Rmd"), but it will be easiest to read the worksheet html file ("Week One.html"). Markdown files are converted to HTML files through a process called "knitting". You can knit the .Rmd file into an HTML file in RStudio using the "Knit" button. To build up your intuition about how markdown files are made, it will be good to flip back and forth between the .Rmd file and the .html file. 

Each week, you will download the .Rmd and .html file associated with that week's lab assignment. We recommend that you immediately save a new version of the .Rmd with your initials appended to the end of the file name. You will edit your own copy of the .Rmd file, filling in answers (typically as chunks of code) under each of the exercises. When you see red, bold text that says "Exercise", there is a practice question to answer. When you have completed the worksheet, you will knit your completed .Rmd file, __check that your file looks correct and that all of your answers are easily read__, and then submit both your final .Rmd file and knitted .html file on Canvas. 

### <span style="color: #E74C3C;">Exercise 1</span>

Flip back to the .Rmd file and identify where you see the following formating syntax: 

- text bookmarked with two underscores
- text in square brackets followed by text in parentheses
- text followed by hashtags

Identify what each of these formatting operations do. Feel free to use the internet if you cannot figure it out, but please do note when these formatting tricks come up in your .Rmd files. 

<br>

## Mathematical Operations

In its most basic form, R can be used as a calculator. Simply type into the console some mathematical expression and hit enter to receive an output. For example:

```{r math 1}
2+2
```

This works just as well with more complicated expressions:

```{r math 2}
(2+(1/3))*4^2
```

Many mathematical operators that we will use require the use of functions that are built into R (see intro to functions below). Here are a few examples, but you will likely need to search for operators when new ones come up. 

* square root: sqrt(x)
* logarithm: log(x, base)
* exponential: exp(x)

Here are some examples:

```{r math 3}
sqrt(36) # square root of 36
log(100, 10) # log base 10 of 100
log(100) # natural log of 100 (when base argument is undefined, the default is e)
exp(2) # e^2
```

<br>

Notice in the chunk above that there is text to the write of the hashtag symbol. This is called a comment, and it is very very important. Within chunks (or within an R Script), everything will be interpreted as code to be run UNLESS it is preceeded by a hashtag. This enables us to make code slightly more readable (though does not have full flexibility of markdown syntax within a markdown file). It is very important that you comment your code so that others (most often a future version of yourself, but also Jeremy) can understand what you're trying to do. Some people like to comment out the goal of some section of code while others like to comment out a description of what the code is doing. I often use both sorts of comments. 

### <span style="color: #E74C3C;">Exercise 2</span>
__If code is required to complete an exercise, please write the code in a chunk. Ensure that the chunk of code and the output of the code is visible in the knitted html document.__

<br>

Write a line of code to calculate $log_636$

<br>


## Storing Information as Objects

Most of the time, we're not just using R as a calculator. Instead, we perform more complicated tasks on sets of data. Data (as well as everything else) is stored in the form of an _object_ that we can then perform tasks on. To store information as an object, we can use the assignment operators "<-" or "=". For example, to store a single value, we can use the following code:

```{r storing 1}
dat.pt1<- 5 # store the value 5 in an object called 'dat.pt1'
dat.pt1
```

We can also store a set of data as an object called a vector. To do this, we may use the "c" function, which stands for concatenate. If you are unfamiliar with this word, the definition will help you understand what the "c" function does. If you are still stumped, type into your console "?c", which will pull up the help page for this function. Here are some examples of storing sets of data as vectors:

```{r storing 2}
var1 <- c(5, 4, 6, 3, 5, 5) # store this vector as an object called "var1"
var2 <- c(8, 9, 9, 9, 8, 7)
var3 <- c("red", "red", "red", "blue", "blue", "blue")
var1
```

Note that we can put all sorts of values in a vector. For example, var1 and var2 are both numeric vectors, while var3 is a character vector. As you code more, you will find it helpful to keep track of what sorts of values are stored within your objects. 

We can also combine vectors into a single dataframe. When we do this, we may wish to treat each vector as a column, in which case we'll want to use the function "cbind.data.frame". Alternatively, we may wish to treat each vector as a row, in which case we'll want to use the function "rbind.data.frame". Often, we have vectors of corresponding values that need to be binded as columns such that each row is a different "observation" with associated values. As an example, let's bind the vectors above into a dataframe:

```{r storing 3}
dat <- cbind.data.frame(var1, var2, var3) # combine these 3 vectors as columbs in a dataframe
dat
```

<br>

There are a few more object types that you will likely come across in this class, including functions (see below), matrices, arrays, and lists. We'll try to explain these object types as we encounter them. 

Quite often, we wish to access the contents of an object. How you access the contents depends on the type of object, and there are often multiple methods for accessing contents. Square brackets are perhaps the most flexible method for accessing the contents of R objects. We can use integer values to index an object and return the specified element. For example, if I want to access the third element of the var1 vector, I can use the following code: 

```{r index1}
var1[3]
```

Similarly, if I wish to access the 3rd column of the dat dataframe, I can use the following code: 

```{r index2}
dat[,3]
```

Indexing two-dimensional objects using square brackets requires two values, the row and column indices, separated by a comma. When one or both values are left blank (as the row value is in the preceeding chunk), all values for that dimension are returned. 

For objects with named elements, we can use the $ operator. For example, the columns of dat are named, so we can rewrite the last chunk as: 

```{r index3}
dat$var3
```

which returns the same values. If we wanted to, we could name the values of a vector, matrix, or list and using the $ operator to retrieve species values too. 

Indexes need not be just one integer or one element name. If we place a vector of integers or names into the square brackets following an object's name, we can return each of the indicated values from the object. Alternatively, we can place a vector of integers or names into the square brackets, immediately following a ! operator, to return all elements EXCEPT the specified elements. 

### <span style="color: #E74C3C;">Exercise 3</span>

Write some code to return the 2nd and 4th value of var2. 

<br>

Finally, we can use Boolean (true or false) to index values. The output is the elements indicated by the position of TRUE values. We can do so explicitly, like in the following chunk of code: 

```{r index4}
var1[c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)]
```

We can use Boolean as indexes via logical statements. Logical statements are evaluated by R and return Boolean, which become the indices if used within square brackets. For example, if I wanted to return the values of var1 for which var3 is "red", I can use the following chunk of code: 

```{r index5}
var1[var3 == "red"]
```

### <span style="color: #E74C3C;">Exercise 4</span>

Create a new dataframe which containing columns for your top 5 favorite organisms, which kingdom each belongs to, and your personal rating of how cute they are on a scale from 1 to 5 with 1 being not cute at all 5 being the cutest thing imaginable (thus, three columns). Then, write a line of code to return the names of any species that scored less than a 3 on cuteness. 

<br>

## Functions and Packages

As you've seen with some of the mathematical operations we'll be using, R contains many useful functions to accomplish a variety of tasks. Functions are objects that do stuff, and they look like a string of characters followed by a set of parentheses. Within the parentheses are arguments that modify what exactly the function does. 

Functions are used to simplify tasks. For example, we could find the sum of all the data for var1 this way:

```{r functions 1}
5 + 4 + 6 + 3 + 5 + 5
```

Or, we could find the sum using the sum() function:

```{r functions 2}
sum(var1)
```

If we were to be working from the dat dataframe, we could do the same thing by specifying var1 within dat like this:

```{r functions 3}
sum(dat$var1)
```

If we wanted to calculate the sum of var1 for rows in which var3 is red, we will need to subset the data. There are a few ways to do this, but here is one:

```{r functions 4}
sum(dat$var1[dat$var3 == "red"])
```

Here, the "==" operator means "is equal to". Note that "=" is already used as an assignment operator (synonymous to "<-"), so we must use "==" when subsetting.  

We can get this same value from taking the sum for var1 in which var3 is NOT blue (because, in this case, if the value is not blue, it is red):

```{r functions 5}
sum(dat$var1[dat$var3 != "blue"])
```

Another note on troubleshooting: you can learn more about a function by Googling it, or by typing in the a "?" followed by the name of the function in the console (for example, type into the console: `?sum`). The function documentation should open in the help tab, and you can find out what the function does, what package it is contained in, and which arguments it takes. 

<br>



### <span style="color: #E74C3C;">Exercise 3</span>
Find the average of var1 for values where var3 is red.

<br>



### <span style="color: #E74C3C;">Exercise 4</span>
Besides the vector to be summed, what additional argument does the sum() function contain? What is the purpose of this additional argument?

<br>


## Generating and Reading in Data

As you've seen, we can generate our own data by typing out individual values. If we want these values as a vector, we use the c() function. Values can be numeric or strings of characters. For example:

```{r generating 1}
size <- c(6.7, 8.7, 12.3, 6.5, 4.9)
id <- c("aa1d", "ab1e", "ac1d", "bb1d", "aa2e")
```

We can also generate our own data using a few operators and functions. For example, if we'd like a sequence of numbers, one way we can do this is:

```{r generating 2}
1:100
```

We can also use the seq() function to be more specific about how we want R to generate a sequence. (try typing in `?seq` to see how it works)

```{r generating 3}
seq(from = 1, to = 100, by = 2)
```

If we want the same value repeated, we can use the rep() function as such:

```{r generating 4}
rep(3, 10)
```

We will learn about other ways to generate data next week when we learn about taking random samples. 

Often, we want to work from raw data without inputting all of the data in R. We can input data from a variety of file types using a set of functions. Most often, we will read in csv files using the read.csv() function and specifying the file path. For example, lets load in a sample dataset. Make sure that you have downloaded the "nutrient_data" file. 

A good habit to get into when reading in data from a source location is to set a working directory. This tells R where to look for files when you try to call them. Then, you can more easily read in the file. In the chunk below, there is an example of what a working directory might look like, but _you will need to edit the chunk to represent the location in which you stored your data._

```{r reading 1}
setwd("~/Documents/GitHub/UO-PopEcol/Data") #setting the working directory
dat <- read.csv("nutrient_data") # reading in the file data and storing as a dataframe
head(dat) # head() outputs the first x lines of a dataframe (default is 6)
```



## Packages

Not all functions that exist are available in base R. Often, we want to use functions that are found within certain packages. Thus, installing and loading packages is an important skill. One package many R users find indispensable is the tidyverse package. To install the tidyverse package, _run this line of code in your console_:

install.packages("tidyverse")

Note that this code is not in a chunk. We will likely compile the markdown file multiple times. Markdown will get finicky about trying to compile a .Rmd file in which we're installing a package. You should only have to install any given package once on your computer. Every time you open R or RStudio, however, you will need to load the package like this:

```{r packages 1, warning = FALSE, message = FALSE}
library(tidyverse)
```



## Data Visualization

Data visualization is an important step for performing many quantitative tasks. Often, we visualize data at various stages during data analysis or model construction. In R, people mostly make plots to visualize their data with the tools already available in 'base R', or with the set of functions in the ggplot2 package (already loaded in the tidyverse package). Here, we'll show how to make a quick scatter plot using both approaches. 


### Scatter plot in base R

Making a scatter plot in Base R is pretty straight forward. We can use the plot() function and define an x and y variable. 

```{r plotting 1}
plot(dat$stem.height, dat$fruit.number)
```

<br>




#### <span style="color: #E74C3C;">Exercise 5</span>
A) Identify and describe at least 5 additional arguments that can be defined to adjust the aesthetics of this plot. *(Hint: start by looking at help file for `plot` but also look at `par` which is mentioned in the plot help)*

B) Make a new plot using these 5+ modifications

<br>


### Scatter plot in ggplot

Plotting in ggplot follows a set of grammatical rules ("gg" stands for "grammar of graphics"). The data to be plotted is established first, and then specifications about how this data should be plotted follows. 

```{r plotting 2}
ggplot(data = dat, aes(x = stem.height, y = fruit.number)) + geom_point()
```

<br>



#### <span style="color: #E74C3C;">Exercise 6</span>
Recreate each of these two plots, but change the axis labels to "Stem Height" and "Number of Fruits per Plant".

<br>


## Practice Problems

This data frame contains fruit count for the study species, but we would like to have data for fecundity as total seed output. From a separate study, we have data for the average seeds per fruit of this species under the three nutrient treatments. That study found that this species produces 3.4, 4.2, and 5.0 seeds per fruit, on average, in low, medium, and high nutrient level treatments, respectively. 

<br>



### <span style="color: #E74C3C;">Exercise 7</span>
Use these averages to create a new variable within the dataframe for the approximate seed output of each plant.

<br>
```{r }
dat <- cbind.data.frame(nutrient = c("l", "l", "m", "m", "h", "h"), 
                 fruit = c(2, 3, 2, 5, 5, 4))
low.ave = 3.4
med.ave = 4.2
high.ave = 5
dat$seed <- ifelse(dat$nutrient == "l", dat$fruit*low.ave, 
                   ifelse(dat$nutrient == "m", dat$fruit*med.ave, dat$fruit*high.ave))
dat$seed
```

### <span style="color: #E74C3C;">Exercise 8</span>
Calculate the average seed output for each nutrient treatment.

<br>



### <span style="color: #E74C3C;">Exercise 9</span>
Make a bar plot of seed output under each nutrient treatment.

<br>



### <span style="color: #E74C3C;">Exercise 10</span>
Make a scatter plot to compare seed output and stem height.

<br>



### <span style="color: #E74C3C;">Exercise 11</span>
Make a scatter plot to compare seed output and stem height, but include a best fit line for each of the three nutrient treatments. Color the points and lines by treatment.

<br>


## Additional resources

We encourage you to continue to practice these skills and learn more about performing tasks in R. If you need additional help with any of these skills or are looking for additional practice with the basics of R, here are some resources:

https://r4ds.had.co.nz/r-markdown.html#introduction-18
http://swcarpentry.github.io/r-novice-inflammation/
https://swcarpentry.github.io/r-novice-gapminder/
https://espm-288.carlboettiger.info/syllabus/
https://www.carlboettiger.info/teaching
https://github.com/laurenmh/bio-data-course-2018

Additionally, here are some helpful pages for going deeper into some of these basics:

https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf [Cheat Sheet for R Markdwon]
https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf [Cheat Sheet for ggplot]
https://towardsdatascience.com/base-plotting-in-r-eb365da06b22 [blog post in base R plotting]
https://r4ds.had.co.nz/introduction.html [using R with the tidyverse]
