---
title: "Introduction to R"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    # keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The goal for this handout is to familiarize ourselves with R, RStudio, and RMarkdown and to gain intuition about how coding can be used in the field of population ecology. We will work through this markdown file, running lines of code and writing our own code to practice basic skills in R. 

Here is a list of learning objectives that we will cover in this worksheet. We will use these skills throughout the course, so it will make your life easier if you get these down early. 

* Basics of troubleshooting (AKA how to use Google)
* Performing basic mathematical operations
* Using functions
* Installing packages
* Generating data
* Reading in data
* Basics of data wrangling
* Using RMarkdown



## Troubleshooting

Likely the most important skill to have under your belt for coding in R is how to troubleshoot. Regardless of your experience level with R, you are bound to run into issues in the form of error messages or undesirable outputs. Luckily, the internet exists. Not only does the internet exist, but the internet contains, within it, practically endless resources regarding R. All one has to do is copy and paste an error message into a search engine or Google "[x topic] in r" and they're bound to find forums and blog posts dedicated to their issue. We encourage you to practice this skill by attempting to find answers through these resources when you run into issues. However, do not waste lots of time on troubleshooting. Seek out help via peers or your course instructor or TA if you're still having trouble after searching for a solution on your own. 



## R Markdown

Markdown files are a compilation of text and lines of code. They're great for communicating the output of code (or the code itself) without the reader getting bogged down in a messy script. There are lots of ways to customize R Markdown files, and there is always more to learn about them, but for now, we'll stick to the basics. What you're reading right now is basic text. We can put in R code by creating a "chunk". While completing this worksheet, you will need to type into the markdown file itself ("Week One.Rmd"), but it will be easiest to read the worksheet html file ("Week One.html"). Markdown files are converted to HTML files through a process called "knitting". You can knit the .Rmd file into an HTML file in RStudio using the "Knit" button. 

To build up your intuition about how markdown files are made, it will be good to flip back and forth between the .Rmd file and the .html file. When you see a large question mark and bold text in the HTML file (or text in the Rmd file bookmarked by "__"), there is a practice question to answer. Please complete this question in the .Rmd file. When you've completed the worksheet, please save a new copy of the .Rmd file with your initials and knit a new HTML file. Upload both files to canvas. 



## Mathematical Operations

In its most basic form, R can be used as a calculator. Simply type into the console some mathematical expression and hit enter to receive an output. For example:

```{r math 1}
2+2
```

This works just as well with more complicated expressions:

```{r math 2}
(2+(1/3))*4^2
```

Many mathematical operators that we will use require the use of functions that are built into R (see intro to functions below). Here are a few examples, but you will likely need to search for operators when new ones come up. 

* square root: sqrt(x)
* logarithm: log(x, base)
* exponential: exp(x)

Here are some examples:

```{r math 3}
sqrt(36) #square root of 36
log(100, 10) #log base 10 of 100
log(100) # natural log of 100 (when base argument is undefined, the default is e)
exp(2) #e^2
```

<br>


### <span style="color: #871FBF;">Exercise 1</span>
__If code is required to complete an exercise, please write the code in a chunk. Ensure that the chunk of code and the output of the code is visible in the knitted html document.__

<br>

Write a line of code to calculate $log_636$

<br>


## Storing Information as Objects

Most of the time, we're not just using R as a calculator. Instead, we perform more complicated tasks on sets of data. We store data in the form of an object that we can then perform tasks on. To store information as an object, we can use the assignment operators "<-" or "=". For example, to store a single value, we can use the following code:

```{r storing 1}
dat.pt1<- 5 # store the value 5 in an object called 'dat.pt1'
dat.pt1
```

We can also store a set of data as an object called a vector. To do this, we may use the "c" function, which stands for concatenate. If you are unfamiliar with this word, the definition will help you understand what the "c" function does. If you are still stumped, type into your console "?c", which will pull up the help page for this function. Here are some examples of storing sets of data as vectors:

```{r storing 2}
var1 <- c(5, 4, 6, 3, 5, 5) # store this vector as an object called "var1"
var2 <- c(8, 9, 9, 9, 8, 7)
var3 <- c("red", "red", "red", "blue", "blue", "blue")
var1
```

We can also combine vectors into a single dataframe. When we do this, we may wish to treat each vector as a column, in which case we'll want to use the function "cbind.data.frame". Alternatively, we may wish to treat each vector as a row, in which case we'll want to use the function "rbind.data.frame". Often, we have vectors of corresponding values that need to be binded as columns such that each row is a different "observation" with associated values. As an example, let's bind the vectors above into a dataframe:

```{r storing 3}
dat <- cbind.data.frame(var1, var2, var3) # combine these 3 vectors as columbs in a dataframe
dat
```

<br>




### <span style="color: #871FBF;">Exercise 2</span>
Run the following lines of code and see what they return. Describe what the $ and [] operators are doing in this code and how to use them. Try to figure it out by examining the output, but if you get stumped, type into the console "?'\$'".

dat$var1

dat$var3

dat$var3[1]

dat$var3[5]

dat[1, 1]

dat[1, 3]

dat[1, ]

dat[, 3]

<br>


## Using Functions

As you've seen with some of the mathematical operations we'll be using, R contains many useful functions to accomplish a variety of tasks. Functions are a string of characters followed by a set of parentheses. Within the parentheses are arguments that specify what you want the function to do. 

Functions are used to simplify tasks. For example, we could find the sum of all the data for var1 this way:

```{r functions 1}
5 + 4 + 6 + 3 + 5 + 5
```

Or, we could find the sum using the sum() function:

```{r functions 2}
sum(var1)
```

If we were to be working from the dat dataframe, we could do the same thing by specifying var1 within dat like this:

```{r functions 3}
sum(dat$var1)
```

If we wanted to calculate the sum of var1 for rows in which var3 is red, we will need to subset the data. There are a few ways to do this, but here is one:

```{r functions 4}
sum(dat$var1[dat$var3 == "red"])
```

Here, the "==" operator means "is equal to". Note that "=" is already used as an assignment operator (synonymous to "<-"), so we must use "==" when subsetting.  

We can get this same value from taking the sum for var1 in which var3 is NOT blue (because, in this case, if the value is not blue, it is red):

```{r functions 5}
sum(dat$var1[dat$var3 != "blue"])
```

Another note on troubleshooting: you can learn more about a function by Googling it, or by typing in the a "?" followed by the name of the function in the console (for example, type into the console: `?sum`). The function documentation should open in the help tab, and you can find out what the function does, what package it is contained in, and which arguments it takes. 

<br>



### <span style="color: #871FBF;">Exercise 3</span>
Find the average of var1 for values where var3 is red.

<br>



### <span style="color: #871FBF;">Exercise 4</span>
Besides the vector to be summed, what additional argument does the sum() function contain? What is the purpose of this additional argument?

<br>


## Generating and Reading in Data

As you've seen, we can generate our own data by typing out individual values. If we want these values as a vector, we use the c() function. Values can be numeric or strings of characters. For example:

```{r generating 1}
size <- c(6.7, 8.7, 12.3, 6.5, 4.9)
id <- c("aa1d", "ab1e", "ac1d", "bb1d", "aa2e")
```

We can also generate our own data using a few operators and functions. For example, if we'd like a sequence of numbers, one way we can do this is:

```{r generating 2}
1:100
```

We can also use the seq() function to be more specific about how we want R to generate a sequence. (try typing in `?seq` to see how it works)

```{r generating 3}
seq(from = 1, to = 100, by = 2)
```

If we want the same value repeated, we can use the rep() function as such:

```{r generating 4}
rep(3, 10)
```

We will learn about other ways to generate data next week when we learn about taking random samples. 

Often, we want to work from raw data without inputting all of the data in R. We can input data from a variety of file types using a set of functions. Most often, we will read in csv files using the read.csv() function and specifying the file path. For example, lets load in a sample dataset. Make sure that you have downloaded the "nutrient_data" file. 

A good habit to get into when reading in data from a source location is to set a working directory. This tells R where to look for files when you try to call them. Then, you can more easily read in the file. In the chunk below, there is an example of what a working directory might look like, but _you will need to edit the chunk to represent the location in which you stored your data._

```{r reading 1}
setwd("~/Documents/GitHub/PopEcol22/S24/Data") #setting the working directory
dat <- read.csv("nutrient_data") # reading in the file data and storing as a dataframe
head(dat) # head() outputs the first x lines of a dataframe (default is 6)
```



## Packages

Not all functions that exist are available in base R. Often, we want to use functions that are found within certain packages. Thus, installing and loading packages is an important skill. One package many R users find indispensable is the tidyverse package. To install the tidyverse package, _run this line of code in your console_:

install.packages("tidyverse")

Note that this code is not in a chunk. We will likely compile the markdown file multiple times. Markdown will get finicky about trying to compile a .Rmd file in which we're installing a package. You should only have to install any given package once on your computer. Every time you open R or RStudio, however, you will need to load the package like this:

```{r packages 1, warning = FALSE, message = FALSE}
library(tidyverse)
```



## Data Visualization

Data visualization is an important step for performing many quantitative tasks. Often, we visualize data at various stages during data analysis or model construction. In R, people mostly make plots to visualize their data with the tools already available in 'base R', or with the set of functions in the ggplot2 package (already loaded in the tidyverse package). Here, we'll show how to make a quick scatter plot using both approaches. 


### Scatter plot in base R

Making a scatter plot in Base R is pretty straight forward. We can use the plot() function and define an x and y variable. 

```{r plotting 1}
plot(dat$stem.height, dat$fruit.number)
```

<br>




#### <span style="color: #871FBF;">Exercise 5</span>
A) Identify and describe at least 5 additional arguments that can be defined to adjust the aesthetics of this plot. *(Hint: start by looking at help file for `plot` but also look at `par` which is mentioned in the plot help)*

B) Make a new plot using these 5+ modifications

<br>


### Scatter plot in ggplot

Plotting in ggplot follows a set of grammatical rules ("gg" stands for "grammar of graphics"). The data to be plotted is established first, and then specifications about how this data should be plotted follows. 

```{r plotting 2}
ggplot(data = dat, aes(x = stem.height, y = fruit.number)) + geom_point()
```

<br>



#### <span style="color: #871FBF;">Exercise 6</span>
Recreate each of these two plots, but change the axis labels to "Stem Height" and "Number of Fruits per Plant".

<br>


## Practice Problems

This data frame contains fruit count for the study species, but we would like to have data for fecundity as total seed output. From a separate study, we have data for the average seeds per fruit of this species under the three nutrient treatments. That study found that this species produces 3.4, 4.2, and 5.0 seeds per fruit, on average, in low, medium, and high nutrient level treatments, respectively. 

<br>



### <span style="color: #871FBF;">Exercise 7</span>
Use these averages to create a new variable within the dataframe for the approximate seed output of each plant.

<br>
```{r }
dat <- cbind.data.frame(nutrient = c("l", "l", "m", "m", "h", "h"), 
                 fruit = c(2, 3, 2, 5, 5, 4))
low.ave = 3.4
med.ave = 4.2
high.ave = 5
dat$seed <- ifelse(dat$nutrient == "l", dat$fruit*low.ave, 
                   ifelse(dat$nutrient == "m", dat$fruit*med.ave, dat$fruit*high.ave))
dat$seed
```

### <span style="color: #871FBF;">Exercise 8</span>
Calculate the average seed output for each nutrient treatment.

<br>



### <span style="color: #871FBF;">Exercise 9</span>
Make a bar plot of seed output under each nutrient treatment.

<br>



### <span style="color: #871FBF;">Exercise 10</span>
Make a scatter plot to compare seed output and stem height.

<br>



### <span style="color: #871FBF;">Exercise 11</span>
Make a scatter plot to compare seed output and stem height, but include a best fit line for each of the three nutrient treatments. Color the points and lines by treatment.

<br>


## Additional resources

We encourage you to continue to practice these skills and learn more about performing tasks in R. If you need additional help with any of these skills or are looking for additional practice with the basics of R, here are some resources:

https://r4ds.had.co.nz/r-markdown.html#introduction-18
http://swcarpentry.github.io/r-novice-inflammation/
https://swcarpentry.github.io/r-novice-gapminder/
https://espm-288.carlboettiger.info/syllabus/
https://www.carlboettiger.info/teaching
https://github.com/laurenmh/bio-data-course-2018

Additionally, here are some helpful pages for going deeper into some of these basics:

https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf [Cheat Sheet for R Markdwon]
https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf [Cheat Sheet for ggplot]
https://towardsdatascience.com/base-plotting-in-r-eb365da06b22 [blog post in base R plotting]
https://r4ds.had.co.nz/introduction.html [using R with the tidyverse]
